---
title: 'Homework: Numerical linear algebra'
output:
  html_document:
    df_print: paged
  html_notebook:
    code_folding: none
    highlight: textmate
---

<script type="text/x-mathjax-config">
MathJax.Hub.Config({
  TeX: {
    Macros: {
      bA: "{\\boldsymbol{A}}",
      bx: "{\\boldsymbol{x}}",
      bb: "{\\boldsymbol{b}}"
    }
  }
});
</script>


```{r setup, include=FALSE}
source(file.path("..", "R", "util.R"))

required_packages <- c("microbenchmark")
install_and_load_packages(required_packages)
```


# Exercise 1: Comparing different numerical linear algebra algorithms for solving linear systems

In this exercise, we consider the problem of solving a linear system $\bA \bx = \bb$ for $\bx$.
We compare the three methods we learned in the class: LU, Cholesky, and QR decompositions.
(Of course, LU applies to more general systems and QR can be used to solve least squares, but here we focus on positive definite systems.)

## Part A: Racing for solution &mdash; speed comparison 

We first compare their computational speed. 
Fill in the code below and `bench::mark()` the three algorithms.

**Questions:**
What are relative speeds among the algorithms?
Do relative speeds agree with what you expect from the complexity analyses?
If not (quite), why might that be?

**Answer:**

The median time for LU, Cholesky, and QR decomposition are each 142ms, 147ms, and 448ms. 
Here, time complexity for each decomposition is $\frac{1}{3}p^3$, $\frac{2}{3}p^3$, and $2np^2-\frac{2}{3}p^3 = \frac{2}{3}p^3$, ($\because n == p$). The relative speeds of Cholesky and QR decomposition roughly agree with what I expect from time complexity. But, the speeds of Cholesky and LU composition are roughly the same as each other where the theoretical speed of Cholesky is half of the speed of LU. 

**Note:**
I misspoke about the R's built-in `chol()` function during the lecture:
when applied to a positive-definite `A`, the function actually returns an _upper-triangular_ matrix `R` such that `t(R) %*% R == A`.

```{r, echo=FALSE}
# Import the `rand_positive_def_matrix` function
source(file.path("R", "random_matrix.R"))

mat_size <- 1024L
A <- rand_positive_def_matrix(mat_size, cond_num = 1e3)
b <- rnorm(mat_size)

solve_via_cholesky <- function(A, b) {
  R = chol(A)
  return(backsolve(R, forwardsolve(t(R), b)))
}

bench::mark(
  solve(A, b)
)
bench::mark(
  solve_via_cholesky(A, b)
)
bench::mark(
  qr.solve(A, b)
)
```


## Part B: Competition in terms of numerical accuracy/stability

We now compare the three methods in terms of numerical accuracy/stability.
To this end, we set up the following simulation study. 
We first generate a "ground truth" solution vector $\bx_0$.
We then compute an "numerical-error-free" $\bb = \bA \bx_0$ by carrying out the matrix-vector multiplication using the `long double` type, which (on most hardware and compilers) provides [additional 12 bits of precision](https://en.wikipedia.org/wiki/Extended_precision#x86_extended_precision_format).
Of course, the vector $\bb$ computed as such still suffers from numerical errors, but the idea is that the numerical errors from this high-precision matrix-vector multiplication is much smaller than the errors caused by numerically solving $\bA \bx = \bb$ for $\bx$.
We can thus assess the accuracy of the three solvers by comparing the numerically computed $\bx$ to the ground truth $\bx_0$.

### Task &#x2F00;

First compare the outputs of matrix-vector multiplication $\bx \to \bA \bx$ using `double` and `long double` using the provided Rcpp functions.

**Questions:**
What is the relative difference in $\ell^2$-norm? 
How about the coordinate-wise relative differences?
Are the observed magnitudes of the differences what you'd expect?

**Answer:** 
The relative difference in $\ell^2$-norm of `double` from `long double` is 5.667e-16 and the mean of the coordinate-wise relative difference of `double` from `long double` is 9.5793-16. The maximum of the coordinate-wise relative difference is 9.514e-13. 

Since magnitudes of relative difference are on the order of $\epsilon_{machine}\kappa(A)$ and the given condition number for A is $1e3$, we can calculate the expected order of the relative errors. 

Machine precision for `double` and `long double` is $2^{-52}$ and $2^{-63}$. Thus, expected order of relative errors are $2^{-52} * 10^3 = 2.220e-13$ and $2^{-63} * 10^3 = 1.084e-16$, respectively. Observed differences roughly agree with expected(theoretical) ones.  

```{r, echo==FALSE}
Rcpp::sourceCpp(file.path("src", "matvec_double.cpp"))
Rcpp::sourceCpp(file.path("src", "matvec_ldouble.cpp"))

set.seed(1918)

mat_size <- 1024L
A <- rand_positive_def_matrix(mat_size, cond_num = 1e3)
x <- rnorm(mat_size)

double_Ax = matvec_double(A, x)
ldouble_Ax = matvec_ldouble(A, x)
rel_diff_coord = (ldouble_Ax - double_Ax) / ldouble_Ax
diff_l2 = sqrt(sum((ldouble_Ax - double_Ax)^2)) 
ldouble_l2 = sqrt(sum((ldouble_Ax)^2))
rel_diff_l2 = diff_l2 / ldouble_l2 # L2-norm rel diff
summary(rel_diff_coord) # coordinate-wise rel diff
```


### Task &#x2F06;

Now randomly generate $\bA$ so that its condition number is $10^6$.
Then solve a positive-definite system $\bA \bx = \bb$ for $\bx$ using the three algorithms and compare their outputs to the ground truth $\bx_0$.

**Questions:**
Which algorithm appears to be more accurate than the others? 
Visually demonstrate your answer.

**Answer:**

The mean value of coordinate-wise relative errors is 1.022e-12, 1.102e-11, and 5.266e-13 for LU, Cholesky, and QR decomposition, respectively. This indicates that QR is the most accurate while Cholesky is the worst. 

```{r warning = FALSE}
set.seed(1918)
cond_num <- 1e6

# Larger matrices could incur substantial computational time under base R BLAS
mat_size <- 1024L 

A <- rand_positive_def_matrix(mat_size, cond_num)
x <- rnorm(mat_size)

l2_norm <- function(x){sqrt(sum(x^2))}
  
b <- matvec_ldouble(A, x)

LU_x <- solve(A, b)
chol_x <- solve_via_cholesky(A, b)
QR_x <- qr.solve(A, b)

relative_LU <- (LU_x - x) / LU_x
relative_chol <- (chol_x - x) / chol_x
relative_QR <- (QR_x - x) / QR_x

library(ggplot2)
library(tidyverse)

relative_coord = data.frame(relative_coord = c(relative_LU, relative_chol, relative_QR), method = rep(c("LU", "chol", "QR"), each = mat_size), index = 1:mat_size)

relative_coord %>% ggplot(aes(index, relative_coord)) +
  geom_line(aes(colour = method))

cat(mean(relative_LU), sd(relative_LU))
cat(mean(relative_chol), sd(relative_chol))
cat(mean(relative_QR), sd(relative_QR))

```


### Task &#x4E09;

In Task &#x2F06;, we compared the three algorithms in one randomly generated example.
Now we consider a more systematic (though hardly comprehensive) comparison via repeated simulations.
We also vary the condition number of $\bA$ and assess whether the results would hold across varying degrees of ill-conditioning.

**Questions/To-do's:**

* Using the starter code provided, calculate various summary measures of the numerical errors.
* Integrate into the provided code one another (or more, if you like) meaningful metric(s) of your choice to summarize the numerical error.
* Visually explore how the three algorithms compare with each other in their accuracy. See if different error metrics tell different stories; they might or might not.
* Vary the condition number in the range $10^6 \sim 10^{12}$, e.g. by trying $10^6$, $10^9$, and $10^{12}$.
* Do you see any patterns in the numerical errors across the three algorithms, metrics, and/or condition numbers? Show some plots to support your conclusion.

**Note:** 
The QR solver will throw an error when the system is ill-condition enough that the numerical solution might not be very accurate. 
To force it to return the solution in any case, set `tol = .Machine$double.eps`.


```{r echo = FALSE}
# Utility functions for bookkeeping simulation results.
source(file.path("R", "num_linalg_sim_study_helper.R"))

numerical_errors <- function(cond_num = 1e6) {
  n_sim <- 250L
  mat_size <- 512L
  metrics <- c("l2_norm", "l1_norm", "median", "five_percentile", "ninety_five_percentile")
  rel_error_list <- lapply( c("lu", "chol", "qr"), 
                            function(method) pre_allocate_error_list(n_sim, metrics))
  
  for (sim_index in 1:n_sim) {
    A <- rand_positive_def_matrix(mat_size, cond_num)
    x <- rnorm(mat_size)
    b <- matvec_ldouble(A, x)
    
    x_approx <- list(
      lu = solve(A, b),
      chol = solve_via_cholesky(A, b),
      qr = qr.solve(A, b, tol = .Machine$double.eps)
    )
    for (method in c("lu", "chol", "qr")) {
      rel_error <- lapply(metrics,
                          function (metric)
                            calc_rel_error(x, x_approx[[method]], metric))
      names(rel_error) <- metrics
      for (metric in names(rel_error)) {
        rel_error_list[[method]][[metric]][sim_index] <- rel_error[[metric]]
      }
    }
  }
  
  rel_error_df <- data.frame()
  
  for (Method in c("lu", "chol", "qr")) {
    error_df <- data.frame(method = Method, error = rel_error_list[[Method]]) %>%
      pivot_longer(!method, "metric", "rel_error")
    rel_error_df <- rbind(rel_error_df, error_df)
  }
  
  out_plot = rel_error_df %>%
    ggplot(aes(x = rel_error)) + 
    geom_density(aes(group = method, fill = method)) + 
    facet_wrap(~ metric) + 
    scale_y_continuous(breaks = c())
  
  return(out_plot)
}
```

## Visually explore how the three algorithms differs in different condition numbers. 
```{r, cache=TRUE}
numerical_errors(1e6)
numerical_errors(1e9)
numerical_errors(1e12)
```
